\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-04 chapters/bttda.tex
Update bttda formulation

\input{chapters/bttda/tensorviz.tex}
\todo{abbreviations}

\section{Introduction}
\todo{immediately arrive to the novelty or to the point}
\todo{reduce length of the introduction}
\todo{less acronyms}
\todo{put the intuition behind the algorithm (sparse, more flexible, what it
does in one sentence ) in the introduction. Additionally "try to get some
information from the data, continue extracting more information as long as
there is some left}
Brain-computer interfaces (BCIs) have the potential to bypass
defective neural pathways by providing an alternative communication channel
between the brain and an external device.
These interfaces find applications in the development of neuroprosthetics and assistive
technologies, among other applications~\cite{Wolpaw2020}.
To achieve their functionality, BCIs record and process neural data obtained through
a neuroimaging technique, with electroencephalography (EEG) being the most popular.
\todo{explicitly state how BTD is a generalization of PARAFAC and TUCKER}

Due to its multichannel time series structure, EEG data, like most neural
signal acquisition modalities used for BCIs, naturally exist as multiway data,
capturing information in both spatial and temporal domains.
Common preprocessing transformations, such as time-frequency transformation,
time-binning, or integrating information across multiple subjects or conditions,
can further expand the data into additional analytic domains.
This can result in high-dimensional datasets which are usually flattened into a
set of sample vectors, stripping the original data from its structure.
Yet, the intrinsic multiway structure of neural data~\cite{Erol2022} is
well-suited for representation as \emph{tensors}, or multiway arrays, where
each domain is represented as a tensor \emph{mode}.
Tensors provide a structured data representation for this highly dimensional
multiway data
This in turn paves the way to the development of tensor methods which can
counteract some of the drawbacks of the dimensionality problem.
Tensor methods are machine learning techniques that  consider each tensor mode
separately, reducing a given problem into partial, per-mode problems.
This approach has given rise to efficient dimensionality reduction techniques,
such as the Higher Order Singular Value Decomposition
(HOSVD)~\cite{DeLathauwer2000,SoleCasals2018}
and Canonical Polyadic Decomposition
(CPD)~\cite{Hitchcock1927,Nazarpour2006}.

HOSVD and CPD are two examples of unsupervised techniques with applications in EEG
processing.
However, given the low signal-to-noise ratio and
specific, task-related output expected in a BCI application,
supervised feature extraction and machine learning techniques are
favored\todo{cite Lotte review}.
%One approach is to incorporate assumptions of the tensor structure of the data
%directly into the estimation of parameters of classic supervised linear -
%machine learning methods, such as linear discriminant analysis (LDA) or beamforming.
%Advances have been made by leveraging the decoupling of the spatial and temporal
%domains in EEG event-related potential (ERP) classification using Spatiotemporal
%Discriminant Analysis~\cite{Li2010,Zhang2013} or methods regularizing covariance
%matrix estimation~\cite{Kerchove2022,Sosulski2022}.

\todo{come in here with François' comment about restructuring the intro.}
A more structured approach to the same problem is to design a supervised
tensor dimensionality reduction method that optimizes discriminability between the
extracted features, as is the case with Higher Order Discriminant
Analysis (HODA)~\cite{Yan2005,Phan2010,Froelich2018}.
Extracted features can subsequently be further classified, most commonly
using LDA or a support vector machine (SVM) to predict class labels.
Variants of HODA have been applied to BCI problems such as
ERP~\cite{Onishi2012,Higashi2016}\todo{cite spatiotemporal discriminant
analysis here, 'variants and special cases'} and motor imagery (MI)~\cite{Liu2015,Cai2021}
decoding.
Recent work proposes adaptations such as suited objective
functions and regularization~\cite{Jamshidi2017,Jorajuria2022,Aghili2023}.

The methods above adhere to the Tucker tensor decomposition
structure,
\todo{
To introduce BTD, I would start with a paragraph explaining the decomposition
itself, or the approximation for dimension reduction, just after the
presentation of tensors: on one side, Tucker-N (1 core only) and on the other
side polyadic decomposition (PD, m scalars) and in the middle the BTD. This
paragraph explains that in most cases the decomposition is not unique and
depends on a priori parameters (nb of cores, dimensions of projection matrices,
etc.) Also defines the residual, difference between the original tensor and the
model in the case of approximation.
Then, the techniques for getting an
"optimal" decomposition or approximation could be presented: HOSVD and equiv.
(just a few refs) in the case of Tucker and CPD and others (just a few refs)
for PD. Then, your state of the art could be just focused on specific
optimizations of BTD used in BCIs (in order to justify the JNE journal
target)
In fact, this is the option chosen by Axel Faes and Marc in the JNE
paper of 2022)...}
meaning that they reduce input tensors of size
$(D_1,D_2,\ldots,D_K)$ to a smaller tensor of size $(r_1,r_2,\ldots,r_K)$ with
each $r_k\leq D_k$, similar to HOSVD.
While effective, other approaches such as the PARAFAC structure employed in
CPD, where a tensor is decomposed into a sum of rank-1 tensors,
might also be suitable to represent the neural data of interest.
Discriminant tensor features can also be extracted
in the PARAFAC structure, for instance through manifold
optimization~\cite{Froelich2018}.

Nevertheless, the PARAFAC structure might still not be able to
efficiently represent all relevant information in a compressed format.
The block-term tensor structure is a generalization of the Tucker and
PARAFAC structures, and can be calculated in an unsupervised way using
the Block-term Tensor Decomposition
(BTD)~\cite{DeLathauwer2008,DeLathauwer2008a,DeLathauwer2008b,Rontogiannis2021}.
BTD, of which the HOSVD and CPD are special cases,
represents a tensor as sum of Tucker terms.
Research has shown that supervised decoders adopting this more flexible structure
can improve BCI performance.
Promising results have been achieved for regression tasks using
Higher Order Parial Least Squares~\cite{Camarrone2018} and Block-Term Tensor
Regression (BTTR)~\cite{Faes2022,Faes2022b}.
BTTR has also been adapted into a classification variant~\cite{Camarrone2021}
but this leaving room for improvement.
Instead of optimizing features directly for class separability, rather a dummy
independent variable was regressed towards, and the method
cannot be extended to a multi-class setting.
Furthermore, structures employed in HOPLS, BTTR are still more constrained than
what could be achieved with a full block-term tensor structured decomposition
optimized for discriminability, since they rely on a low rank common subspace
between the input and classification labels.
\todo{remove acronyms if not used later}

%\todo{Describe spectrum-weighted tda~\cite{Huang2020}, structure is not
%	flexible enough, not broadly applicable, not interpretable}

We propose to mitigate these issues by designing a new supervised feature
extraction tensor method based on the abovementioned HODA algorithm
that is more suited for the extraction of discriminant
features while adhering to a flexible and efficient block-term tensor
structure.
This work features the following contributions:
\begin{enumerate*}[label={\arabic*)}]
	\item  We develop a forward model for HODA to reconstruct a
	      given input tensor from the extracted features.
	\item This allows us to introduce a state-of-the-art BCI feature extraction
        method based on the block-term tensor structure, named Block-Term Tensor Discriminant Analysis
	      (BTTDA).
  \item We evaluate a BCI decoder based on BTTDA and its special
	      PARAFAC-structured case on decoding tasks for both ERP and MI
	      paradigm BCI datasets.
\end{enumerate*}
\todo{It's always trichiin in the related work section of a ML/Mathematical
paper, but unless you r reviewers are absolute experts of this specific field,
you should at least try to better describe the previous methods.}

\section{Methods}

@@

HODA optimizes discriminability in the Fisher sense, by optimizing the
Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$
\begin{equation}
	\phi = \frac{\sum_c^CN_c\left\lVert\bar{\ten{G}}(c)-\bar{\bar{\ten{G}}}\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{G}(n)-\bar{\ten{G}}(c_n)\right\rVert_F^2}
	\label{eq:fisher}
\end{equation}
for $C$ classes with each $N_c$ samples. $\bar{\ten{G}}(c)$ is the mean of
latent
tensors of class $c$, and $\bar{\bar{\ten{G}}}$ the mean of
these class mean latent tensors.
The goal is now to find the optimal projection matrices\todo{You should say
somewhere that the shape (r1,r2,...rK) is fixed a priori, no?}
\begin{equation}
	\left\{\mat{U}^*\right\} = \argmax_{\{\mat{U}\}}\phi
\end{equation}
which can be solved by the iterative algorithm in Algorithm~\ref{alg:HODA}.
\begin{algorithm}
  \caption[The \ac{hoda} backward solution.]{The \ac{hoda} backward solution.}
	\label{alg:HODA}
	\input{chapters/bttda/alg_HODA.tex}
\end{algorithm}
To start, $\mat{U}_k$ are initialized to orthogonal matrices, e.g. as random
orthonormal matrices, by a per mode Singular Value Decomposition (SVD),
or as the partial HOSVD of all stacked tensors in the dataset.
At each iteration, the algorithm loops through the modes and fixes all
projections but $\mat{U}_k$ corresponding to mode $k$.
It then finds a partial latent tensor
\begin{equation}
	\ten{G}_{-k}=\ten{X}\mmprs{\mat{U}}{k}
\end{equation}
Subsequently, a new projection matrix $\mat{V}_k$ can be found analogous to Linear
Discriminant Analysis by constructing the within-class scatter matrix
\begin{equation}
	\mat{S}_{-k,\text{w}} = \sum_n^N\tilde{\mat{G}}_{-k,k}(n)\cdot\tilde{\mat{G}}_{-k,k}^\intercal(n)
\end{equation}
with $\tilde{\ten{G}}_{-k}(n) = \ten{G}_{-k}(n) - \bar{\ten{G}}_{-k}(c_n)$,
and the between-class scatter matrix
\begin{equation}
	\mat{S}_{-k,\text{b}} =
	\sum_c^CN_c\tilde{\bar{\mat{G}}}_{-k,k}(c)\cdot\tilde{\bar{\mat{G}}}_{-k,k}^\intercal(c)
\end{equation}
with $\tilde{\bar{\ten{G}}}_{-k}(c) = \bar{\ten{G}}_{-k}(c) - \bar{\bar{\ten{G}}}_{-k}$,
and solving for the $r_k$ leading eigenvectors in the eigenvalue problem
\begin{equation}
	\mat{S}_{-k,\text{b}}-\varphi_k\mat{S}_{-k,\text{w}} =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
with $\varphi_k=\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{b}}\mat{U}_k\right)/\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{w}}\mat{U}_k\right)$
using the $\mat{U}_k$ obtained in the previous iteration.
Finally, the orthogonal transformation invariant projections $\mat{U}_k$
are obtained by calculating the
per-mode total scatter matrices
\begin{equation}
  \mat{S}_{k,\text{t}} = \sum_n^N\mat{X}_k(n)\cdot\mat{X}_k^\intercal(n)
\end{equation}
and finding the $r_k$ leading eigenvectors of
\begin{equation}
	\mat{V}_k\mat{V}_k^\intercal\mat{S}_{k,\text{t}}\mat{V}_k\mat{V}_k^\intercal
	= \mat{U}_k\mat{\Lambda}\mat{U}_k^\intercal
\end{equation}
at each iteration~\cite{Wang2007}.
%and solving the generalized eigenvalue problem
%\begin{equation}
%  = \mat{\lambda}\mat{S}_{-k,\text{w}}\mat{U}_k
%\end{equation}
The iterative process halts after a fixed amount or iterations, or when the
update of each $\mat{U}_k$ is lower than a predetermined threshold.

@@

To avoid overfitting and improve performance in low sample size settings, the
HODA problem can be regularized by shrinking the partial
within-class scatter matrices~\cite{Phan2010} with a shrinkage factor
$\alpha_k$ at each step such that the eigenvalue problem becomes
\todo{make sure this appears after varphi is defined, move reference to
algorithm to later}
\begin{equation}
	\mat{S}_b^{(-k)} -
	\varphi\left[\left(1-\alpha_k\right)\mat{S}_{-k,\text{w}}+\alpha_k\mat{I}\right] =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
As in Linear Discriminant Analysis, the shrinkage parameter $\alpha_k$ can
also be estimated in a data driven way in HODA~\cite{Jorajuria2022},
e.g., using the Ledoit-Wolf procedure~\cite{Ledoit2003} at every iteration.

@@

The HODA projection in Equation~\ref{eq:hoda-backward}\todo{missing} is an example
of a backward model.
A straightforward and computationally efficient candidate for a corresponding
forward model is given by
\begin{equation}
	\ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}}
	\label{eq:HODA-forward}
\end{equation}
with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$
and error term $\ten{\mathbfcal{E}}$.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_forward.tikz.tex}
  \caption[The \ac{hoda} forward projection.]{The forward projection for HODA. By calculating activation
		patterns $A_k$, the original tensor $\ten{X}$ can approximately be
		reconstructed from projected latent tensor $\ten{G}$. $A_k$ are chosen such
		that the variability captured in the latent tensor is maximally expressed in
		the reconstructed tensor and not in the error term.}
	\label{fig:HODA-forward}
\end{figure}
\todo{not referred to in text}
\todo{I would prefer a figure in which the Ai and G are on the left and
$\hat{X}$on the right. So, rather than using a "almost equal to" sign you could
use a standard equal signal. Finally, in the caption, you could clearly define
the original(X) and reconstructed ($\hat{X}$) tensors as well as the error term.}
\todo{explicitly state why choosing this versus full model or HOPLS}

A forward model should make sure that reconstruction error is minimized and
variation captured in the latent tensor is maximally captured by the forward
projection term $\ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
\todo{not clear ("error is minimized" in the same sentence)}
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\text{E}\left[\text{vec}\left({\ten{\mathbfcal{E}}(n)}\right)\text{vec}\left({\ten{G}(n)}\right)\right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares (ALS) algorithm~\cite{Bentbib2022},%~\cite{Comon2009},
iteratively fixing all but one of the activation patterns such that
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed amount of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in Algorithm~\ref{alg:HODA-fw}.
\begin{algorithm}
  \caption[The \ac{hoda} forward solution.]{The \ac{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}
\todo{Reorder figures and add error term}

@@

Since BTTDA is specified as a forward model, a backward modelling
procedure is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
BTTDA to be useful as a feature extraction method.
The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
computed through a deflation scheme summarized in Algorithm~\ref{alg:BTTDA}.
\begin{algorithm}
  \caption{\Ac{bttda}.}
	\label{alg:BTTDA}
	\input{chapters/bttda/alg_bttda.tex}
\end{algorithm}
For each block $b$, the latent tensor is extracted using the HODA backward
projection in \todo{ref equation}from a residual error term
$\ten{\mathbfcal{E}}^{(b)}$
\begin{equation}
	\ten{G}^{(b)} = \ten{\mathbfcal{E}}^{(b)}\mmpr{\mat{U}^{(b)}}
\end{equation}
This residual error term is calculated by finding the difference between the
previous error and its reconstruction after backward and forward HODA
projection
\begin{equation}
	\ten{\mathbfcal{E}}^{(b+1)} = \ten{\mathbfcal{E}}^{(b)} - \ten{G}^{(b)}
	\mmpr{\mat{A}^{\intercal(b)}}
\end{equation}
with $\ten{\mathbfcal{E}}^{(1)}=\ten{X}$.

@@

\subsection{Model and feature selection}
Similar to the unsupervised BTD, the performance of BTTDA is
heavily dependent on the rank $(r_1^{(b)}, r_2^{(b)}, \ldots,
	r_K^{(b)})$ and on the number of blocks $B$.
If these are not known a priori, i.e. if they cannot be set based on insights in the
data generation process, a model selection step is necessary in order to
determine what the optimal values for $r_k^{(b)}$ and $B$ are.

>>>>

\input{chapters/bttda/tensorviz.tex}

\section{Introduction}
\todo{reduce length of the introduction}

Brain-computer interfaces (BCIs) have the potential to bypass
defective neural pathways by providing an alternative communication channel
between the brain and an external device.
These interfaces find applications in the development of neuroprosthetics and assistive
technologies, among other applications~\cite{Wolpaw2020}.
To achieve their functionality, BCIs record and process neural data obtained through
a neuroimaging technique, with electroencephalography (EEG) being the most popular.

Due to its multichannel time series structure, EEG data, like most neural
signal acquisition modalities used for BCIs, naturally exist as multiway data,
capturing information in both spatial and temporal domains.
Common preprocessing transformations, such as time-frequency transformation,
time-binning, or integrating information across multiple subjects or conditions,
can further expand the data into additional analytic domains.
This can result in high-dimensional datasets which are usually flattened into a
set of sample vectors, stripping the original data from its structure.
Yet, the intrinsic multiway structure of neural data~\cite{Erol2022} is
well-suited for representation as \emph{tensors}, or multiway arrays, where
each domain is represented as a tensor \emph{mode}.
Tensors provide a structured data representation for this highly dimensional
multiway data
This in turn paves the way to the development of tensor methods which can
counteract some of the drawbacks of the dimensionality problem.
Tensor methods are machine learning techniques that  consider each tensor mode
separately, reducing a given problem into partial, per-mode problems.

These tensor methods decompose a tensor into a lower dimensional structure of a
core tensor and factor tensors.
The most common aproaches adhere to either the Tucker structure or the PARAFAC
structure.
A Tucker decomposition reduces an input tensor of size $(D_1,D_2,\ldots,D_K)$ to
a dense  tensor of size $(r_1,r_2,\ldots,r_K)$ with $r_k<=D_k$ using a
set of per mode factor matrices.
Effective unsupervised tensor decomposition and approximation in the Tucker format can be achieved
using the Higher Order Singular Value
Decomposition (HOSVD)~\cite{DeLathauwer2000,SoleCasals2018}.
Alternatively, the PARAFAC structure can be used.
Here, the tensor is decomposed into a sum of rank-1 tensors, each the product
of a scalar and a vector per mode.
This is equivalent to a Tucker structured decomposition with all core elements
off the hyperdiagonal set to 0.
One way of obtaining an unsupervised PARAFAC decomposition is through the Canonical Polyadic
Decomposition~\cite{Hitchcock1927,Nazarpour2006}.

While commonly used, these Tucker or PARAFAC structure might still not be able to
efficiently represent relevant neural information in a compressed format.
The block-term tensor structure is a generalization of the Tucker and
PARAFAC structures.
It represents the tensor as a sum of Tucker structured terms.
If the number of terms is equal to 1, it is equivalent to the Tucker structure;
if the rank of each term is equal to 1, it is equivalent to the PARAFAC
structure.
The generalized structure can be calculated in an unsupervised way using
the Block-Term Decomposition~\cite{DeLathauwer2008,DeLathauwer2008a,DeLathauwer2008b,Rontogiannis2021}.
Performance of methods leveraging either the Tucker and PARAFAC structure are
heavily dependent on the a priori choice of hyperparameters describing the
respectively the multilinear rank or the number of rank-1 terms.
The block-term structure is more flexible than either the Tucker or the PARAFAC
structures, since it is not constrained to problems that cannot be expressed by
one of these structures and the chosen hyperparameters.
However, this increased flexibility also increased the number of
hyperparameters to both the number of terms and the multilinear rank of each term.

If the decompositions are not full rank, the Tucker, PARAFAC and block-term
structures are not unique and can be obtained by optimizing different criteria.
Given the low signal-to-noise ratio and specific, task-related output expected
in a BCI application, supervised feature extraction and machine learning techniques are
favored~\cite{Lotte2018} over the unsupervised decomposition methods presented
above.
A decomposition that is helpful for classification purposes should ideally optimize
the discriminability between classes in the resulting core tensors, which can
be considered as extracted features.
In this philosophy, the Tucker decomposition can also be obtained
using Higher Order Discriminant Analysis
(HODA)~\cite{Yan2005,Phan2010,Froelich2018}, which optimizes class separability in the Fischer sense, analogous to linear
discriminant analysis.
Extracted features can subsequently be further classified, most commonly
using LDA or a support vector machine (SVM) to predict class labels.

Variants of HODA have been applied to BCI problems such as
ERP~\cite{Onishi2012,Higashi2016} and motor imagery (MI)~\cite{Liu2015,Cai2021}
decoding with positive results~\cite{Lotte2018}.
Recent work proposes adaptations such as suited objective
functions and regularization~\cite{Jamshidi2017,Jorajuria2022,Aghili2023}.
Discriminant tensor features have also be extracted
in the PARAFAC structure through manifold optimization~\cite{Froelich2018}.
However, it is not immediately obvious if either the Tucker or PARAFAC
structure are most suited to represent the neural data of interest for the \ac{bci}
paradigm and for decoding.

Recent research has shown that supervised decoders adopting a more flexible structure
can improve BCI performance.
Promising results have been achieved for regression tasks using
Higher Order Parial Least Squares~\cite{Camarrone2018} and Block-Term Tensor
Regression (BTTR)~\cite{Faes2022,Faes2022b}.
BTTR has also been adapted into a classification variant~\cite{Camarrone2021}
but this leaving room for improvement.
Instead of optimizing features directly for class separability, rather a dummy
independent variable was regressed towards, and the method
cannot be extended to a multi-class setting.
Furthermore, structures employed in these regression are still more constrained
than what could be achieved with a full block-term tensor structured decomposition
optimized for discriminability, since they rely on a low rank common subspace
between the input and classification labels.
\textcite{Huang2020} propose a supervized approach for finding multiple discriminant
multilinear spectral filter terms and apply it to motor imagery \ac{bci}, but their
decomposition is also restricted in flexibility.

\todo{It's always trichiin in the related work section of a ML/Mathematical
paper, but unless you r reviewers are absolute experts of this specific field,
you should at least try to better describe the previous methods.}

A block-term decomposition that is directly optimized for discriminability and with a
good choice of ranks and number of terms might better represent the behaviour
of generators of neural activity through increased flexibility, or might
achieve better regularization through increased sparsity.
An alternative view on the same approach goes as follows:
If HODA with a well chosen multilinear rank extracts some discriminant features
from the input tensor, it is likely that it does not yet retrieve all useful
information due to the restriction following from its Tucker structure.
Could we therefore not keep using HODA to extract discriminant Tucker structured
as long as decoding performance increases?

We propose to implement this idea as a new supervised feature
extraction tensor method that is a generalization of the aforementioned HODA
algorithm and is more suited for the extraction of discriminant
features while adhering to a flexible and efficient block-term tensor
structure.
This work features the following contributions:
\begin{enumerate*}[label={\arabic*)}]
	\item  We develop a forward model for HODA to reconstruct a
	      given input tensor from the extracted features.
	\item This allows us to introduce a state-of-the-art BCI feature extraction
        method based on the block-term tensor structure, named Block-Term Tensor Discriminant Analysis
	      (BTTDA).
  \item We evaluate a BCI decoder based on BTTDA and its special
	      PARAFAC-structured case on decoding tasks for both ERP and MI
	      paradigm BCI datasets.
\end{enumerate*}
\section{Methods}

@@

HODA optimizes discriminability in the Fisher sense, by optimizing the
Fisher ratio $\phi$ between the latent tensors $\ten{G}(n)$
\begin{equation}
	\phi = \frac{\sum_c^CN_c\left\lVert\bar{\ten{G}}(c)-\bar{\bar{\ten{G}}}\right\rVert_F^2}
	{\sum_n^N\left\lVert\ten{G}(n)-\bar{\ten{G}}(c_n)\right\rVert_F^2}
	\label{eq:fisher}
\end{equation}
for $C$ classes with each $N_c$ samples. $\bar{\ten{G}}(c)$ is the mean of
latent
tensors of class $c$, and $\bar{\bar{\ten{G}}}$ the mean of
these class mean latent tensors.
If the ranks $(r_1,r_2, \ldots,r_k)$ are set a priori, the goal is now to find the optimal projection matrices
\begin{equation}
	\left\{\mat{U}^*\right\} = \argmax_{\{\mat{U}\}}\phi
\end{equation}
which is solved through the HODA algorithm.
To start, $\mat{U}_k$ are initialized to orthogonal matrices, e.g. as random
orthonormal matrices, by a per mode Singular Value Decomposition (SVD),
or as the partial HOSVD of all stacked tensors in the dataset.
At each iteration, the algorithm loops through the modes and fixes all
projections but $\mat{U}_k$ corresponding to mode $k$.
It then finds a partial latent tensor
\begin{equation}
	\ten{G}_{-k}=\ten{X}\mmprs{\mat{U}}{k}
\end{equation}
Subsequently, a new projection matrix $\mat{V}_k$ can be found analogous to Linear
Discriminant Analysis by constructing the within-class scatter matrix
\begin{equation}
	\mat{S}_{-k,\text{w}} = \sum_n^N\tilde{\mat{G}}_{-k,k}(n)\cdot\tilde{\mat{G}}_{-k,k}^\intercal(n)
\end{equation}
with $\tilde{\ten{G}}_{-k}(n) = \ten{G}_{-k}(n) - \bar{\ten{G}}_{-k}(c_n)$,
and the between-class scatter matrix
\begin{equation}
	\mat{S}_{-k,\text{b}} =
	\sum_c^CN_c\tilde{\bar{\mat{G}}}_{-k,k}(c)\cdot\tilde{\bar{\mat{G}}}_{-k,k}^\intercal(c)
\end{equation}
with $\tilde{\bar{\ten{G}}}_{-k}(c) = \bar{\ten{G}}_{-k}(c) - \bar{\bar{\ten{G}}}_{-k}$,
and solving for the $r_k$ leading eigenvectors in the eigenvalue problem
\begin{equation}
	\mat{S}_{-k,\text{b}}-\varphi_k\mat{S}_{-k,\text{w}} =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
with $\varphi_k=\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{b}}\mat{U}_k\right)/\Tr\left(\mat{U}_k^\intercal\mat{S}_{-k,\text{w}}\mat{U}_k\right)$
using the $\mat{U}_k$ obtained in the previous iteration.
Finally, the orthogonal transformation invariant projections $\mat{U}_k$
are obtained by calculating the
per-mode total scatter matrices
\begin{equation}
  \mat{S}_{k,\text{t}} = \sum_n^N\mat{X}_k(n)\cdot\mat{X}_k^\intercal(n)
\end{equation}
and finding the $r_k$ leading eigenvectors of
\begin{equation}
	\mat{V}_k\mat{V}_k^\intercal\mat{S}_{k,\text{t}}\mat{V}_k\mat{V}_k^\intercal
	= \mat{U}_k\mat{\Lambda}\mat{U}_k^\intercal
\end{equation}
at each iteration~\cite{Wang2007}.
%and solving the generalized eigenvalue problem
%\begin{equation}
%  = \mat{\lambda}\mat{S}_{-k,\text{w}}\mat{U}_k
%\end{equation}
The iterative process halts after a fixed amount or iterations, or when the
update of each $\mat{U}_k$ is lower than a predetermined threshold.
The full HODA procedure is summarized in Algorithm~\ref{alg:HODA}.
\begin{algorithm}
  \caption[The \ac{hoda} backward solution.]{The \ac{hoda} backward solution.}
	\label{alg:HODA}
	\input{chapters/bttda/alg_HODA.tex}
\end{algorithm}

@@

To avoid overfitting and improve performance in low sample size settings, the
HODA problem can be regularized by shrinking the partial
within-class scatter matrices~\cite{Phan2010} with a shrinkage factor
$\alpha_k$ at each step such that the eigenvalue problem becomes
\begin{equation}
	\mat{S}_b^{(-k)} -
	\varphi\left[\left(1-\alpha_k\right)\mat{S}_{-k,\text{w}}+\alpha_k\mat{I}\right] =
	\mat{V}_k\mat{\Lambda}\mat{V}_k^\intercal
\end{equation}
As in Linear Discriminant Analysis, the shrinkage parameter $\alpha_k$ can
also be estimated in a data driven way in HODA~\cite{Jorajuria2022},
e.g., using the Ledoit-Wolf procedure~\cite{Ledoit2003} at every iteration.

@@

The HODA projection in Equation~\ref{eq:hoda-backward}\todo{missing} is an example
of a backward model.
A straightforward and computationally efficient candidate for a corresponding
forward model is given by
\begin{equation}
  \ten{X} = \ten{G}\mmpr{\mat{A}^\intercal} + \ten{\mathbfcal{E}} =
  \hat{\ten{X}} + \ten{\mathbfcal{E}}
	\label{eq:HODA-forward}
\end{equation}
with \emph{activation patterns} $\mat{A}_k \in \mathbb{R}^{D_k\times r_k}$,
reconstructed tensor $\hat{\ten{X}}$ and error term $\ten{\mathbfcal{E}}$.
\begin{figure}[t]
	\centering
	\input{figures/bttda/HODA_forward.tikz.tex}
  \caption[The \ac{hoda} forward projection.]{The forward projection for HODA.
    By calculating activation	patterns $\mat{A}_k$, the original tensor $\ten{X}$ can approximately be
    reconstructed as $\hat{\ten{X}}$ from projected latent tensor $\ten{G}$.
    The reconstruction is accurate up to an error term $\ten{\mathbfcal{E}}$,
    such that $\ten{X}=\hat{\ten{X}}+\ten{\mathbfcal{E}}$.
    $\mat{A}_k$ are chosen such that the variability captured in the latent tensor is
    maximally expressed in the reconstructed tensor $\hat{\ten{X}}$ and not in
    the error term $\ten{\mathbfcal{E}}$.}
	\label{fig:HODA-forward}
\end{figure}
\todo{not referred to in text}
\todo{I would prefer a figure in which the Ai and G are on the left and
$\hat{X}$on the right. So, rather than using a "almost equal to" sign you could
use a standard equal signal. Finally, in the caption, you could clearly define
the original(X) and reconstructed ($\hat{X}$) tensors as well as the error term.}
\todo{explicitly state why choosing this versus full model or HOPLS}

A forward model should make sure that reconstruction error is minimized and
variation captured in the latent tensor is maximally captured by the
reconstruction term $\hat{\ten{X}}= \ten{G}\mmpr{\mat{A}^\intercal}$, and not by the error term
\todo{not clear ("error is minimized" in the same sentence)}
$\ten{\mathbfcal{E}}$~\cite{Haufe2014}.
Hence, we aim to minimize the expected value of the cross-covariance between
the noise term and the extracted latent tensors
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\text{E}\left[\text{vec}\left({\ten{\mathbfcal{E}}(n)}\right)\text{vec}\left({\ten{G}(n)}\right)\right]_n
\end{equation}
or, equivalently~\cite{Parra2005,Haufe2014},
\begin{equation}
	\left\{\mat{A}^*\right\} = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n)
  -\hat{\ten{X}}(n)\right)^2
  = \arg\min_{\{\mat{A}\}}\sum_n^N\left(\ten{X}(n) - \ten{G}(n)\mmpr{\mat{A}}\right)^2
\end{equation}
This least squares tensor approximation problem can be solved efficiently using the
alternating least squares (ALS) algorithm~\cite{Bentbib2022},%~\cite{Comon2009},
iteratively fixing all but one of the activation patterns such that
\begin{equation}
	\mat{A}_k = \argmin_{\mat{A}_k}
	\sum_n^N\left[\mat{X}_k(n) -
		\mat{A}_k\left(\ten{G}(n)\mmprs{\mat{A}}{k}\right)_k\right]^2
\end{equation}
at every iteration, which can be solved directly by ordinary least squares.
The activation patterns are initialized to the weights $\{\mat{U}\}$ of the
backward model.
Similar to fitting the backward model, the iterative process for the forward
model halts after a fixed amount of iterations or when the update of each
$\mat{A}_k$ is lower than a predetermined threshold.
The full algorithm to determine the HODA forward projection is listed
in Algorithm~\ref{alg:HODA-fw}.
\begin{algorithm}
  \caption[The \ac{hoda} forward solution.]{The \ac{hoda} forward solution.}
	\label{alg:HODA-fw}
	\input{chapters/bttda/alg_HODA_fw.tex}
\end{algorithm}
\todo{Reorder figures and add error term}

@@

Since BTTDA is specified as a forward model, a backward modelling
procedure is required which finds the latent tensors $\ten{G}^{(b)}$ given $\ten{X}$ for
BTTDA to be useful as a feature extraction method.
The extracted features represented by the latent tensors $\ten{G}^{(b)}$ can be
computed through a deflation scheme summarized in Algorithm~\ref{alg:BTTDA}.
\begin{algorithm}
  \caption{\Ac{bttda}.}
	\label{alg:BTTDA}
	\input{chapters/bttda/alg_bttda.tex}
\end{algorithm}
For each block $b$, the latent tensor is extracted using the HODA backward
projection in \todo{ref equation} from the residual error term of the previous
block $\ten{\mathbfcal{E}}^{(b-1)}$
\begin{equation}
	\ten{G}^{(b)} = \ten{\mathbfcal{E}}^{(b-1)}\mmpr{\mat{U}^{(b)}}
\end{equation}
This residual error term is calculated by finding the difference between the
previous error and its reconstruction after backward and forward HODA
projection
\begin{equation}
  \ten{\mathbfcal{E}}^{(b)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \hat{\ten{\mathbfcal{E}}}^{(b-1)}
  = \ten{\mathbfcal{E}}^{(b-1)} - \ten{G}^{(b)}\mmpr{\mat{A}^{\intercal(b)}}
\end{equation}
with $\ten{\mathbfcal{E}}^{(0)}=\ten{X}$.

@@

\subsection{Model and feature selection}
Similar to the unsupervised Block-Term Decomposition, the performance of BTTDA is
heavily dependent on the rank $(r_1^{(b)}, r_2^{(b)}, \ldots,
	r_K^{(b)})$ and on the number of blocks $B$.
If these are not known a priori, i.e. if they cannot be set based on insights in the
data generation process, a model selection step is necessary in order to
determine what the optimal values for $r_k^{(b)}$ and $B$ are.
